Speed and Precision in Range Analysis

### 2012-05-24 doc/Manuscript/SPE/paper.tex
Changed the journal paper a tiny bit.

% The problem: few implementations, not extensive experimental reports.
Given this great importance, it comes as no surprise that the compiler
literature is rich in works describing in details algorithmic variations of
range analyses~\cite{Gawlitza09,Mahlke01,Stephenson00,Su05}.
On the other hand, none of these authors provide experimental evidence that
their approaches are able to deal with very large programs.
There are researchers who have implemented range analyses that scale up to
large programs~\cite{Blanchet03,Patterson95,Venet04}; nevertheless, because the
algorithm itself is not the main focus of their works, they neither give
details about their design choices nor provide experimental data about it.
This scenario was recently changed by Oh {\em et al.}~\cite{Oh12}, who
introduced an abstract interpretation framework which processes programs with
hundreds of thousands of lines of code.
Nevertheless, Oh {\em et al.} have designed a simple range analysis,
which does not handle comparisons between variables, for instance.
They also do not discuss the precision of their implementation, but only its
runtime and memory consumption.
In this paper we claim to push this discussion considerably further.

@@

% Our experimental numbers.
We have implemented our algorithm in the LLVM compiler~\cite{Lattner04}, and
have used it to process a test suite with 2.72 million lines of C code.
As we show in Section~\ref{sub:showdown}, our implementation, which is
publicly available, is able to analyze programs with over one million assembly
instructions within fifteen seconds.
And our implementation is not a straw-man: it produces very precise results.
We have compared the ranges that our implementation finds with the results
obtained via a dynamic profiler, which we have also implemented.
As we show in Section~\ref{sub:showdown}, when analyzing well-known numeric
benchmarks we are able to estimate tight ranges for almost half of all the
integer variables present in these programs.
Our results are similar to Stephenson
{\em et al}'s~\cite{Stephenson00}, even though out analysis does not require
a backward propagation phase.
Furthermore, we have been able to find tight bounds to the majority of the
examples used by Costan {\em et al.}~\cite{Costan05} and Lakhdar
{\em et al.}~\cite{Lakhdar11}, who rely on much more costly methods.

@@

From the lattice $\varset{Z}$ we define the product lattice
$\varset{Z}^2$, which is partially ordered by the subset relation
$\sqsubseteq$.
$\varset{Z}^2$ is defined as follows:

\begin{equation*}
\varset{Z}^2 = \emptyset \cup \{[z_1, z_2] | \ z_1,z_2 \in \varset{Z},
\ z_1 \leq z_2, \  -\infty < z_2 \}
\end{equation*}

Given an interval $\iota = [l, u]$, we let $\lb{\iota} = l$, and
$\ub{\iota} = u$.
We let \varset{V} be a set of constraint variables, and
$I: \varset{V} \mapsto \varset{Z}^2$ a
mapping from these variables to intervals in $\varset{Z}^2$.
Our objective is to solve a constraint system $C$, formed by constraints such
as those seen in Figure~\ref{fig:eval_function}(left).
Figure~\ref{fig:eval_function}(right) defines a valuation function $e$ that
computes $Y = f(\ldots)$, given I.
Armed with these concepts, we define the range analysis problem as follows:

@@

\begin{figure}[hbt]
\begin{center}
\begin{small}
\begin{eqnarray*}
\begin{array}{r@{\hspace{1.5cm}}c}
Y = [l, u]
&
e(Y) = [l, u]
\\
\\
Y = \phi (X_1, X_2)
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2]}
{e(Y) = [\mbox{min}(l_1, l_2), \mbox{max}(u_1, u_2)]}
\\
\\
Y = X_1 + X_2
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2]}
{e(Y) = [l_1 + l_2, u_1 + u_2]}
\\
\\
Y = X_1 \times X_2
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2] \\ L = \{l_1l_2, l_1u_2, u_1l_2, u_1u_2\}}
{e(Y) = [\mbox{min}(L), \mbox{max}(L)]}
\\
\\
Y = aX + b
&
\inferrule{I[X]=[l, u] \\ k_l = al + b \\ k_u = au + b}
{e(Y) = [\mbox{min}(k_l, k_u), \mbox{max}(k_l, k_u)]}
\\
\\
Y = X \sqcap [l', u']
&
\inferrule{I[X]=[l, u]}
{e(Y) \leftarrow [\mbox{max}(l, l'), \mbox{min}(u, u')]}
\end{array}
\end{eqnarray*}
\caption{\label{fig:eval_function}
A suite of constraints that produce an instance of the range analysis problem.}
\end{small}
\end{center}
\end{figure}

We will use the program in Figure~\ref{fig:ex1}(a) as the running example
to illustrate our range analysis.
Figure~\ref{fig:ex1}(b) shows the same program in SSA form~\cite{Cytron91},
and Figure~\ref{fig:ex1}(c) outlines the constraints that we extract from this
program.
There is a clear correspondence between instructions and constraints.
A possible solution to the range analysis problem, as obtained via the
techniques that we will introduce in Section~\ref{sec:algo}, is given in
Figure~\ref{fig:ex1}(d).
The SSA form, so conspicuous in modern compilers, leads to a very conservative
solution.
As we will see shortly, we can improve this solution substantially by using
a more sophisticated program representation -- the e-SSA form -- which
gives us path-sensitiveness.

@@

Given that the exact rules to convert a program to e-SSA form have never been
explicitly stated in the literature, we describe our rules as follows.
Let $(v < c)?$ be a conditional test, and let $l_t$ and $l_f$ be labels in
the program, such that $l_t$ is the target of the test if the condition is true,
and $l_f$ is the target when the condition is false.
We split the live range of $v$ at any of these points if at least one of two
conditions is true:
(i) $l_f$ or $l_t$ dominate any use of $v$;
(ii) there exist a use of $v$ at the dominance frontier of $l_f$ or $l_t$.
For the notions of dominance and dominance-frontier, see Aho
{\em et al.}~\cite[p.656]{Aho06}.
To split the live range of $v$ at $l_f$ we insert at this
program point a copy $v_f = v \sqcap [c, +\infty]$, where $v_f$ is a fresh name.
We then rename every use of $v$ that is dominated by $l_f$ to $v_f$.
Dually, if we must split at $l_t$, then we create at this point a copy
$v_t = v \sqcap [-\infty, c-1]$, and rename variables accordingly.
If the conditional uses two variables, e.g., $(v_1 < v_2)?$, then we create
intersections bound to {\em futures}.
We insert, at $l_f$, $v_{1f} = v_1 \sqcap [\fun{ft}(v_2), +\infty]$,
and $v_{2f} = v_2 \sqcap [-\infty, \fun{ft}(v_1)]$.
Similarly, at $l_t$ we insert
$v_{1v} = v_1 \sqcap [-\infty, \fun{ft}(v_2) - 1]$
and $v_{2f} = v_2 \sqcap [\fun{ft}(v_1) + 1, +\infty]$.

We use the notation $\fun{ft}(v)$ to denote the {\em future} bounds of a
variable.
As we will show in Section~\ref{sub:micro}, once the growth pattern of $v$ is
known, we can replace $\fun{ft}(v)$ by an actual value.
Once we are done placing copies, we insert $\phi$-functions into the
transformed program to convert it to SSA form.
This last step avoids that two different names given to the same original
variable be simultaneously alive at the program code.
Figure~\ref{fig:ex_standard_eSSA}(a) shows our running example changed into
standard e-SSA form.
The part (b) of this figure shows the solution that we get to this new
program.
The e-SSA form allows us to bind interval information directly to the live
ranges of variables; thus, giving us the opportunity to solve range analysis
sparsely.
More traditional approaches, which we call {\em dense analyses}, bind
interval information to pairs formed by variables and program points.

@@

\noindent
\textbf{Strongly Connected Components.}
To solve range analysis we find all the strongly connected components (SCCs) of
the dependence graph and collapse them into single nodes; thus, obtaining a
directed acyclic graph.
We then sort the resulting DAG topologically, and apply the analyses from
Section~\ref{sub:micro} on every SCC in topological order.
Once we solve the range analysis problem for a SCC, we propagate the
intervals that we found to the variable nodes at the {\em frontier} of this
SCC.
A variable node $N_v$ is said to be in the frontier of a strongly connected
component $S$ if:
(i) $N_v \notin S$; and
(ii) there exists a variable node $N_v' \in S$, and a constraint node $N_c$,
such that $N_v \leftarrow N_c$, and $N_c \leftarrow N_v'$.
This propagation ensures that when analyzing a strongly connected component $S$
any influence that $S$ might suffer from nodes outside it has already been
considered.

When finding strongly connected components, we take control dependence
edges into consideration.
For instance, in Figure~\ref{fig:ex_graph} the nodes that correspond to the
variables $i_1$, $i_2$, $i_t$, $j_1$, $j_2$ and $j_t$ form a single component.
The dashed edges, which represent control dependences, keep all these variables
connected.
In this way, we ensure that, upon stumbling upon an interval associated with
future bounds, e.g., $\fun{ft}(v)$, either variable $v$ has been solved in a
previous component, or it belongs to the current component.
In the latter case, as we will see soon, we can still take $v$'s interval into
consideration.
As we show in Section~\ref{sub:showdown}, most of the strong components in
actual programs are singletons.
Furthermore, the composite components tend to be small.
These two facts ensure that the more costly parts of our algorithm only have to
handle small inputs.

@@

\noindent
\textbf{Example:}
Continuing with our example, Figure~\ref{fig:ex_partition_grow_crop} shows
the application of our algorithm on the last strong component of
Figure~\ref{fig:ex_graph}.
Upon meeting this SCC, we have already determined that the interval
$[0, 0]$ is bound to $I_0$ and that the interval $[100, 100]$ is bound to
$J_0$, as we show in Figure~\ref{fig:ex_graph}(a).
We are not guaranteed to find the least fix point of a constraint system.
However, in this example we did it.
We emphasize that finding this tight solution was only possible because of
the topological ordering of the constraint graph in
Figure~\ref{fig:ex_graph}.
Had we applied the widening operator onto the whole graph, then we would
have found out that variable $j_0$ is bound to $[-\infty, +\infty]$,
because
(i) it receives its interval directly from variable $k_t$, which is upper
bounded by $+\infty$, and
(ii) it is part of a negative cycle.
On the other hand, because we only analyze $j$'s SCC after we have
analyzed $k$'s, $k$ only contribute the constant range $[0, 99]$ to $j_0$.

@@

Figure~\ref{fig:wholeImpact} compares the precision of the intra and
inter-procedural analyses for the five largest programs in three different
categories of benchmarks: SPEC CPU 2006, the Stanford Suite~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}} and
Bitwise~\cite{Stephenson00}.
Our results for the SPEC programs were disappointingly: on the average for
the five largest programs, the intra-procedural version of our analysis saves
5.23\% of bits per variable.
The inter-procedural version increases this number to 8.89\%.
A manual inspection of the SPEC programs reveals that this result is expected:
these programs manipulate files, and their source codes do no provide enough
explicit constants to power our analysis up.
However, with numerical benchmarks we fare much better.
On the average our inter-procedural algorithm reduces the bitwidth of the
Stanford benchmarks by 36.24\%.
For Bitwise we obtain a bitwidth reduction of 12.27\%.
However, this average is lowered by two outliers: \texttt{edge\_detect} and
\texttt{sha}, which cannot be reduced.
The bitwise benchmarks were implemented by Stephenson
{\em et al.}~\cite{Stephenson00} to validate their bitwidth analysis.
Our results are on par with those found by the original authors.
The bitwise programs contain only the \texttt{main} function; thus, different
versions of our algorithm find the same results when applied onto these
programs.

\subsection{Context Sensitive versus Context Insensitive Analysis}
\label{sub:context}

Another way to increase the precision of range analysis is via a
context-sensitive implementation.
Context-sensitiveness allows us to distinguish different calling sites of
the same function.
Figure~\ref{fig:context} shows why the ability to make this distinction is
important for precision.
In Figure~\ref{fig:context}(a) we have two different calls of function
\texttt{foo}.
An usual way to perform a data-flow analysis inter-procedurally is to
create assignments between formal and actual parameters, as we show in
Figure~\ref{fig:context}(b).
If a function is called more than once, then its formal parameters will receive
information from many actual parameters.
We use $\phi$-functions to bind this information together into a single
flow.
However, in this case the multiple assignment of values to parameters
makes the ranges of these parameters very large, whereas in reality they
are not.
A way to circumvent this source of imprecision is via function inlining,
as we show in Figure~\ref{fig:context}(c).
The results that we can derive for the transformed program are more
precise, as each input parameter is assigned a single value.

@@

\noindent
\textbf{Intra vs Inter-procedural runtimes.}
Figure~\ref{fig:timeComp}(Right) compares three different execution modes.
Bars are normalized to the time to run the intra-procedural analysis
without inlining.
On the average, the intra-procedural mode is 28.92\% faster than the
inter-procedural mode.
If we perform function inlining, then this difference is 45.87\%.
These numbers are close because our runtime is bound to the
size of the strong components.
We have observed that function inlining does not increase too much these
components.

@@

% Scalability: Astree and others.
There have been many practical approaches to abstract interpretation,
with special emphasis on range analysis~\cite{Bertrane10,Blanchet03,Cousot09,Gampe11,Jung05}.
Cousot's group, for instance, has been able to globally analyze programs with
thousands of lines of code, albeit they use domain specific tools.
Astr\'{e}, for example, analyzes programs that do not contain recursive calls.
The work that is the closest to us is Oh {\em et al.}'s very recent abstract
interpretation framework~\cite{Oh12}.
Oh {\em et al.} discuss an implementation of range analysis on the interval
lattice that scales up to a program with 1,363K LoC (ghostscript-9.00), but
they do not provide results about its precision.
We could not find the benchmarks used in those experiments for a
direct comparison -- the distribution of ghostscript-9.00 available in the
LLVM test suite has 27K LoC.
On the other hand, we globally analyzed our largest benchmark, SPEC CPU 2006's
\texttt{gcc}, enabling function inlining, in less than 15 seconds.
Oh {\em et al.}'s implementation took orders of magnitude more time to go over
programs of similar size.
However, whereas they provide a framework to develop general sparse analyses, we
only solve range analysis on the interval lattice.

>>>>

% The problem: few implementations, not extensive experimental reports.
Given this great importance, it comes as no surprise that the compiler
literature is rich in works describing in details algorithmic variations of
range analyses~\cite{Gawlitza09,Mahlke01,Stephenson00,Su05}.
On the other hand, none of these authors provide experimental evidence that
their approaches are able to deal with very large programs.
There are researchers who have implemented range analyses that scale up to
large programs~\cite{Blanchet03,Patterson95,Venet04}; nevertheless, because the
algorithm itself is not the main focus of their works, they neither give
details about their design choices nor provide experimental data about it.
This scenario was recently changed by Oh {\em et al.}~\cite{Oh12}, who
introduced an abstract interpretation framework which processes programs with
hundreds of thousands of lines of code.
Nevertheless, Oh {\em et al.} have designed a simple range analysis,
which does not handle comparisons between variables, for instance.
They do not discuss the precision of their implementation, but only its
runtime and memory consumption.
In this paper we claim to push this discussion considerably further.

@@

% Our experimental numbers.
We have implemented our algorithm in the LLVM compiler~\cite{Lattner04}, and
have used it to process a test suite with 2.72 million lines of C code.
As we show in Section~\ref{sub:showdown}, our implementation, which is
publicly available, is able to analyze programs with over one million assembly
instructions within fifteen seconds.
And our implementation is not a straw-man: it produces very precise results.
We have compared the ranges that our implementation finds with the results
obtained via a dynamic profiler, which we have also implemented.
As we show in Section~\ref{sub:showdown}, when analyzing well-known numeric
benchmarks we are able to estimate tight ranges for almost half of all the
integer variables present in these programs.
Our results are similar to Stephenson
{\em et al}'s~\cite{Stephenson00}, even though our analysis does not require
a backward propagation phase.
Furthermore, we have been able to find tight bounds to the majority of the
examples used by Costan {\em et al.}~\cite{Costan05} and Lakhdar
{\em et al.}~\cite{Lakhdar11}, who rely on much more costly methods.

@@

From the lattice $\varset{Z}$ we define the product lattice
$\varset{Z}^2$, which is defined as follows:
%
\begin{equation*}
\varset{Z}^2 = \{ \emptyset \} \cup \{[z_1, z_2] | \ z_1,z_2 \in \varset{Z},
\ z_1 \leq z_2, \  -\infty < z_2 \}
\end{equation*}
%
This interval lattice is partially ordered by the subset relation, which we
denote by ``$\sqsubseteq$".
The meet operator ``$\sqcap$" is defined by:
\[
[a_1, a_2] \sqcap [b_1, b_2] =
\begin{cases}
[\mbox{max}(a_1, b_1), \mbox{min}(a_2, b_2)], \ \mbox{if} \ a_1 \leq b_1 \leq a_2 \ \mbox{or} \ b_1 \leq a_1 \leq b_2 \\
[a_1, a_2] \sqcap [b_1, b_2] = \emptyset, \ \mbox{otherwise}
\end{cases}
\]

Given an interval $\iota = [l, u]$, we let $\lb{\iota} = l$, and
$\ub{\iota} = u$.
We let \varset{V} be a set of constraint variables, and
$I: \varset{V} \mapsto \varset{Z}^2$ a
mapping from these variables to intervals in $\varset{Z}^2$.
Our objective is to solve a constraint system $C$, formed by constraints such
as those seen in Figure~\ref{fig:eval_function}(left).
We let the $\phi$-functions be as defined by Cytron
{\em et al.}~\cite{Cytron91}: they join different variable names into a single
definition.
Figure~\ref{fig:eval_function}(right) defines a valuation function $e$ on the
interval domain.
Armed with these concepts, we define the range analysis problem as follows:

@@

\begin{figure}[t!]
\begin{center}
\begin{small}
\begin{eqnarray*}
\begin{array}{r@{\hspace{1.5cm}}c}
Y = [l, u]
&
e(Y) = [l, u]
\\
\\
Y = \phi (X_1, X_2)
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2]}
{e(Y) = [\mbox{min}(l_1, l_2), \mbox{max}(u_1, u_2)]}
\\
\\
Y = X_1 + X_2
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2]}
{e(Y) = [l_1 + l_2, u_1 + u_2]}
\\
\\
Y = X_1 \times X_2
&
\inferrule{I[X_1]=[l_1, u_1] \\ I[X_2]=[l_2, u_2] \\ L = \{l_1l_2, l_1u_2, u_1l_2, u_1u_2\}}
{e(Y) = [\mbox{min}(L), \mbox{max}(L)]}
\\
\\
Y = aX + b
&
\inferrule{I[X]=[l, u] \\ k_l = al + b \\ k_u = au + b}
{e(Y) = [\mbox{min}(k_l, k_u), \mbox{max}(k_l, k_u)]}
\\
\\
Y = X \sqcap [l', u']
&
\inferrule{I[X]=[l, u]}
{e(Y) \leftarrow [\mbox{max}(l, l'), \mbox{min}(u, u')]}
\end{array}
\end{eqnarray*}
\caption{\label{fig:eval_function}
A suite of constraints that produce an instance of the range analysis problem.}
\end{small}
\end{center}
\end{figure}

We will use the program in Figure~\ref{fig:ex1}(a) as the running example
to illustrate our range analysis.
Figure~\ref{fig:ex1}(b) shows the same program in SSA form~\cite{Cytron91},
and Figure~\ref{fig:ex1}(c) outlines the constraints that we extract from this
program.
There is a clear correspondence between instructions and constraints.
A possible solution to the range analysis problem, as obtained via the
techniques that we will introduce in Section~\ref{sec:algo}, is given in
Figure~\ref{fig:ex1}(d).
The SSA form, so common in modern compilers, leads to a very conservative
solution.
As we will see shortly, we can improve this solution substantially by using
a more sophisticated program representation -- the e-SSA form -- which
gives us path-sensitiveness.

@@

Given that the exact rules to convert a program to e-SSA form have never been
explicitly stated in the literature, we describe our rules as follows.
Let $(v < c)?$ be a conditional test, and let $l_t$ and $l_f$ be labels in
the program, such that $l_t$ is the target of the test if the condition is true,
and $l_f$ is the target when the condition is false.
We split the live range of $v$ at any of these points if at least one of two
conditions is true:
(i) $l_f$ or $l_t$ dominate any use of $v$;
(ii) there exist a use of $v$ at the dominance frontier of $l_f$ or $l_t$.
For the notions of dominance and dominance-frontier, see Aho
{\em et al.}~\cite[p.656]{Aho06}.
To split the live range of $v$ at $l_f$ we insert at this
program point a copy $v_f = v \sqcap [c, +\infty]$, where $v_f$ is a fresh name.
We then rename every use of $v$ that is dominated by $l_f$ to $v_f$.
Dually, if we must split at $l_t$, then we create at this point a copy
$v_t = v \sqcap [-\infty, c-1]$, and rename variables accordingly.
If the conditional uses two variables, e.g., $(v_1 < v_2)?$, then we create
intersections bound to {\em futures}.
We insert, at $l_f$, $v_{1f} = v_1 \sqcap [\fun{ft}(v_2), +\infty]$,
and $v_{2f} = v_2 \sqcap [-\infty, \fun{ft}(v_1)]$.
Similarly, at $l_t$ we insert
$v_{1v} = v_1 \sqcap [-\infty, \fun{ft}(v_2) - 1]$
and $v_{2f} = v_2 \sqcap [\fun{ft}(v_1) + 1, +\infty]$.
Notice that a variable $v$ can never be associated with a future bound to
itself, e.g., $\fun{ft}(v)$.
This invariant holds because whenever the e-SSA conversion associates a variable
$u$ with $\fun{ft}(v)$, then $u$ is a fresh name created to split the live range
of $v$, given that $v$ was used in a conditional.

We use the notation $\fun{ft}(v)$ to denote the {\em future} bounds of a
variable.
As we will show in Section~\ref{sub:micro}, once the growth pattern of $v$ is
known, we can replace $\fun{ft}(v)$ by an actual value.
After splitting the live ranges according to the rules stated above, we might
have to insert $\phi$-functions into the transformed program to re-convert it to
SSA form.
This last step avoids that two different names given to the same original
variable be simultaneously alive at the program code.
Figure~\ref{fig:ex_standard_eSSA}(a) shows our running example changed into
standard e-SSA form.
We have not created variable names for $i_f$ and $j_f$, because neither
$i_1$ nor $i_j$ are dominated by the target of the conditional's else case.
In this example, new $\phi$-functions are not necessary: new variable
names are not alive together with the original variables.
The part (b) of this figure shows the solution that we get to this new
program.
The e-SSA form allows us to bind interval information directly to the live
ranges of variables; thus, giving us the opportunity to solve range analysis
sparsely.
More traditional approaches, which we call {\em dense analyses}, bind
interval information to pairs formed by variables and program points.

@@

\noindent
\textbf{Strongly Connected Components.}
To solve range analysis we find all the strongly connected components (SCCs) of
the dependence graph and collapse them into single nodes, obtaining a
directed acyclic graph.
We then sort the resulting DAG topologically, and apply the analyses from
Section~\ref{sub:micro} on every SCC in topological order.
Once we solve the range analysis problem for a SCC, we propagate the
intervals that we found to the variable nodes at the {\em frontier} of this
SCC.
A variable node $N_v$ is said to be in the frontier of a strongly connected
component $S$ if:
(i) $N_v \notin S$; and
(ii) there exists a variable node $N_v' \in S$, and a constraint node $N_c$,
such that $N_v \leftarrow N_c$, and $N_c \leftarrow N_v'$.
This propagation ensures that when analyzing a strongly connected component $S$
any influence that $S$ might suffer from nodes outside it has already been
considered.

When finding strongly connected components, we take control dependence
edges into consideration.
For instance, in Figure~\ref{fig:ex_graph} the nodes that correspond to the
variables $i_1$, $i_2$, $i_t$, $j_1$, $j_2$ and $j_t$ form a single component.
The dashed edges, which represent control dependences, keep all these variables
connected.
In this way, we ensure that, upon stumbling upon an interval associated with
future bounds, e.g., $\fun{ft}(v)$, either variable $v$ has been solved in a
previous component, or it belongs to the current component.
In the latter case, as we will see soon, we can still take $v$'s interval into
consideration.
This flexibility is possible because we first discover how each variable in
a strong component grows, before resolving future bounds.
As we show in Section~\ref{sub:showdown}, most of the strong components in
actual programs are singletons.
Furthermore, the composite components tend to be small.
These two facts ensure that the more costly parts of our algorithm only have to
handle small inputs.

An interesting question is: can a variable $v$ be associated with a future
bound of itself, e.g., $\fun{ft}(v)$?
The answer is yes, but only if the program contains pathological comparisons
between a variable and itself, e.g., $v < v$, or $v == v$.
All these comparisons

@@

\noindent
\textbf{Example:}
Continuing with our example, Figure~\ref{fig:ex_partition_grow_crop} shows
the application of our algorithm on the last strong component of
Figure~\ref{fig:ex_graph}.
Upon meeting this SCC, we have already determined that the interval
$[0, 0]$ is bound to $I_0$ and that the interval $[0, 99]$ is bound to
$J_0$, as we show in Figure~\ref{fig:ex_graph}(a).
We are not guaranteed to find the least fix point of a constraint system.
However, in this example we did it.
We emphasize that finding this tight solution was only possible because of
the topological ordering of the constraint graph in
Figure~\ref{fig:ex_graph}.
Had we applied the widening operator onto the whole graph, then we would
have found out that variable $j_0$ is bound to $[-\infty, +\infty]$,
because
(i) it receives its interval directly from variable $k_t$, which is upper
bounded by $+\infty$, and
(ii) it is part of a negative cycle.
On the other hand, because we only analyze $j$'s SCC after we have
analyzed $k$'s, $k$ only contribute the constant range $[0, 99]$ to $j_0$.

@@

Figure~\ref{fig:wholeImpact} compares the precision of the intra and
inter-procedural analyses for the five largest programs in three different
categories of benchmarks: SPEC CPU 2006, the Stanford Suite~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}} and
Bitwise~\cite{Stephenson00}.
Our results for the SPEC programs were disappointingly: on the average for
the five largest programs, the intra-procedural version of our analysis saves
5.23\% of bits per variable.
The inter-procedural version increases this number to 8.89\%.
A manual inspection of the SPEC programs reveals that this result is expected:
these programs manipulate files, and their source codes do no provide enough
explicit constants to power our analysis up.
However, with numerical benchmarks we fare much better.
On the average our inter-procedural algorithm reduces the bitwidth of the
Stanford benchmarks by 36.24\%.
For Bitwise we obtain a bitwidth reduction of 12.27\%.
However, this average is lowered by two outliers: \texttt{edge\_detect} and
\texttt{sha}, which have been purposely engineered to be resilient against
range analyses~\cite{Stephenson00}.
The bitwise benchmarks were implemented by Stephenson
{\em et al.}~\cite{Stephenson00} to validate their intra-procedural bitwidth
analysis.
Our results are on par with those found by the original authors.
The bitwise programs contain only the \texttt{main} function; thus, different
versions of our algorithm find the same results when applied onto these
programs.

\subsection{Achieving Partial Context-Sensitiveness via Function Inlining}
\label{sub:context}

Another way to increase the precision of range analysis is via a
context-sensitive implementation.
Context-sensitiveness allows us to distinguish different calling sites of
the same function.
Figure~\ref{fig:context} shows why the ability to make this distinction is
important for precision.
In Figure~\ref{fig:context}(a) we have two different calls of function
\texttt{foo}.
If we apply the trivial intra-procedural approach of Section~\ref{sub:whole},
then we get the graph shown in Figure~\ref{fig:context}(b).
In other words, if a function is called more than once, then its formal
parameters will receive information from many actual parameters.
We use $\phi$-functions to bind this information together into a single
flow.
However, in this case the multiple assignment of values to parameters
makes the ranges of these parameters very large, whereas in reality they
are not.
A way to circumvent this source of imprecision is via function inlining,
as we show in Figure~\ref{fig:context}(c).
The results that we can derive for the transformed program are more
precise, as each input parameter is assigned a single value.

@@

\noindent
\textbf{Intra vs Inter-procedural runtimes.}
Figure~\ref{fig:timeComp}(Right) compares three different execution modes.
Bars are normalized to the time to run the intra-procedural analysis
without inlining.
On the average, the intra-procedural mode is 28.92\% faster than the
inter-procedural mode.
If we perform function inlining, then this difference is 45.87\%.
These numbers are close because our runtime is bound to the
size of the strong components.
Although function inlining can increase the number of strongly connected
components in the constraint graph, it cannot increase the size of the largest
component, when compared to the simple inter-procedural analysis of
Section~\ref{sub:whole}.

@@

% Scalability: Astree and others.
There have been many practical approaches to abstract interpretation,
with special emphasis on range analysis~\cite{Bertrane10,Blanchet03,Cousot09,Gampe11,Jung05}.
Cousot's group, for instance, has been able to globally analyze programs with
thousands of lines of code, albeit they use domain specific tools.
Astr\'{e}e, for example, analyzes programs that do not contain recursive calls.
The work that is the closest to us is Oh {\em et al.}'s very recent abstract
interpretation framework~\cite{Oh12}.
Oh {\em et al.} discuss an implementation of range analysis on the interval
lattice that scales up to a program with 1,363K LoC (ghostscript-9.00), but
they do not provide results about its precision.
We could not find the benchmarks used in those experiments for a
direct comparison -- the distribution of ghostscript-9.00 available in the
LLVM test suite has 27K LoC.
On the other hand, we globally analyzed our largest benchmark, SPEC CPU 2006's
\texttt{gcc}, enabling function inlining, in less than 15 seconds.
Oh {\em et al.}'s implementation took orders of magnitude more time to go over
programs of similar size.
However, whereas they provide a framework to develop general sparse analyses, we
only solve range analysis on the interval lattice.
